Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_wetland.Rmd
more prallel-comit-issues fixed

# NDVI - Wetland ecosystems  {#NDVI-indicator-wetland}

@@

|Region            |V1    |V3    |V4   |V6   | V9   |V10|
|------------------|------|------|-----|-----|------|---|
|Northern Norway   |638   |275   |55   |102  | 360  |315|
|Central Norway    |702   |1250  |34   |0    | 302  |116|
|Western Norway    |296   |111   |30   |0    | 460  |158|
|Eastern Norway    |251   |89    |103  |0    | 251  |223|
|Southern Norway   |58    |44    |5    |0    | 155  |102|

In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region.

@@

**Where are the wetland sites?**
```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.wetland) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
    labels = c(
      "V1",
      "V3",
      "V4",
      "V6",
      "V9",
      "V10"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

# How are polygon sizes distributed?
```{r histogram-sentinall-ndviw, fig.cap="Histogram of wetland ecosystem polygon sizes in Norway."}
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.

**How does NDVI vary over the years**
```{r NDVI-values-over-time-ndviw, fig.cap="A figure showing NDVI values for wetland ecosystem over time for six different main nature types in varying degrees of condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r NDVI-major-ndviw, warning=FALSE, fig.cap="NDVI values in major wetland ecosystem types in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows"}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")
```
```{r NDVI-basic-ndviw, warning=FALSE, fig.cap="NDVI values in basic wetland ecosystem types in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

**How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="DVI variation across regions in wetland ecosystems (major types) in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

There is no obvious trends in geography.

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)**
```{r ndvi-condition-ndviw, warning=FALSE, fig.cap="NDVI values across wetland sites in different ecological condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
  facet_wrap(~hovedtype)
```

In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, 
but in minertrophic fens (V1) there is not much of a pattern. 
Overlaps between condition classes are generally very large.

**Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
```{r important-variables-ndviwv9, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
nin.wetland2 %>%
  filter(hovedtype == "V9") %>%
  filter(NiN_variable_code %in% c("1AG-B", "7GR-GI", "1AG-A-G", "1AR-C-L", "7TK", "PRSL")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "1AG-B" = "Shrub cover",
    "7GR-GI" = "Abundance of ditches",
    "1AG-A-G" = "Tree cover",
    "1AR-C-L" = "Woody plants in field layer",
    "7TK" = "Heavy vehicles",
    "PRSL" = "Wear/tear"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in semi-natural mires (V9)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
```
```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
nin.wetland2 %>%
  filter(hovedtype == "V10") %>%
  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7JB-GJ")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "7JB-BA" = "Land use intensity",
    "7RA-SJ" = "Succession",
    "7FA" = "Alien species",
    "7JB-GJ" = "Fertilization"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 2) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
```

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. ", fig.height=12}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
  facet_grid(region ~ hovedtype)
```

@@

**Northern Norway**
```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Northern Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.N)$coefficients
```

@@

**Central Norway**
```{r central-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Central Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

@@

**Western Norway**
```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for fou wetland ecosystems in W-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.W)$coefficients

@@

**Eastern Norway**
```{r east-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in E-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.E)$coefficients
```

@@

**Southern Norway**
```{r south-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in S-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.S)$coefficients
```

@@

Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
# plot
allSatNDVI.wetland %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  facet_wrap(~hovedtype, ncol = 1)
```

@@

- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition. This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
- How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.

The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF). Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above. At the same time, wetlands are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.

@@

>>>>

# NDVI - wetland ecosystems  {#NDVI-indicator-wetland}

NDVI (greenness) indicator, wetland ecosystems

@@

| Region          | V1  | V3   | V4  | V6  | V9  | V10 |
|-----------------|-----|------|-----|-----|-----|-----|
| Northern Norway | 638 | 275  | 55  | 102 | 360 | 315 |
| Central Norway  | 702 | 1250 | 34  | 0   | 302 | 116 |
| Western Norway  | 296 | 111  | 30  | 0   | 460 | 158 |
| Eastern Norway  | 251 | 89   | 103 | 0   | 251 | 223 |
| Southern Norway | 58  | 44   | 5   | 0   | 155 | 102 |

In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region. (These polygons likely lie in water and thus outside of the regional polygons due to imprecise coordinates from the mapping device).

@@

**Where are the wetland sites?**

```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE, alpha=0) +
  tm_borders() +
  tm_shape(SentinelNDVI.wetland) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
    labels = c(
      "V1",
      "V3",
      "V4",
      "V6",
      "V9",
      "V10"
    ),
    title = "Major ecosystem type",
    size = 1.5
  ) +
  tm_layout(
    main.title = "Wetland ecosystems \n(NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**
```{r histogram-sentinall-ndviw, fig.cap="Histogram of wetland ecosystem polygon sizes in Norway."}
summary(SentinelNDVI.wetland$area_meters)
hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
abline(v = 100, lty = 2)
```

Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.

**How does NDVI vary over the years?**
```{r NDVI-values-over-time-ndviw, fig.cap="A figure showing NDVI values for wetland ecosystem over time for six different main nature types in varying degrees of condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Year")
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r NDVI-major-ndviw, warning=FALSE, fig.cap="NDVI values in major wetland ecosystem types in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows"}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  ggtitle("NDVI for different ecosystem types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")

```{r NDVI-basic-ndviw, warning=FALSE, fig.cap="NDVI values in basic wetland ecosystem types in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("NDVI for different ecosystem sub-types in good condition") +
  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
```

@@

**How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="DVI variation across regions in wetland ecosystems (major types) in good condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("NDVI for different ecosystems in good condition across regions") +
  labs(y = "NDVI values (Sentinel-2)", x = "Region")
```
There is no obvious trends in geography, with the exception of V3 which shows lower values in Central Norway.

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)**
```{r ndvi-condition-ndviw, warning=FALSE, fig.cap="NDVI values across wetland sites in different ecological condition. V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
SentinelNDVI.wetland %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
  ggtitle("NDVI for different ecosystems in different condition classes") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```
In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, 
but in minertrophic fens (V1) there is not much of a pattern. 
Overlaps between condition classes are generally very large.

**Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
```{r important-variables-ndviwv9, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
nin.wetland2 %>%
  filter(hovedtype == "V9") %>%
  filter(NiN_variable_code %in% c("1AG-B", "7GR-GI", "1AG-A-G", "1AR-C-L", "7TK", "PRSL")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "1AG-B" = "Shrub cover",
    "7GR-GI" = "Abundance of ditches",
    "1AG-A-G" = "Tree cover",
    "1AR-C-L" = "Woody plants in field layer",
    "7TK" = "Heavy vehicles",
    "PRSL" = "Wear/tear"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in semi-natural mires (V9)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
```

```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
nin.wetland2 %>%
  filter(hovedtype == "V10") %>%
  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7JB-GJ")) %>%
  mutate(NiN_variable_des = recode(NiN_variable_code,
    "7JB-BA" = "Land use intensity",
    "7RA-SJ" = "Succession",
    "7FA" = "Alien species",
    "7JB-GJ" = "Fertilization"
  )) %>%
  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
  geom_point(size = 2) +
  geom_jitter(width = 0.25, height = 0.25) +
  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 2) +
  theme(legend.position = "none") +
  ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
```

@@

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. ", fig.height=12}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.wetland %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
```

@@

**Northern Norway**
```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Northern Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Northern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.N)$coefficients
```

@@

**Central Norway**
```{r central-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Central Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Central Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```

@@

**Western Norway**
```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in W-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Western Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Western Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.W)$coefficients

@@

**Eastern Norway**
```{r east-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in E-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Eastern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Eastern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.E)$coefficients
```

@@

**Southern Norway**
```{r south-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in S-Norway."}
SentinelNDVI.wetland.train %>%
  filter(region == "Southern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 1, shape = 16, color = "grey", position = position_jitter(.1) ) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
  ggtitle("Southern Norway") +
  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
  facet_wrap(~hovedtype)
```
```{r}
summary(model.wetland.cond.Sent.S)$coefficients
```

@@

Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
# plot
allSatNDVI.wetland %>%
  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
  geom_point() +
  ggtitle("NDVI across time") +
  labs(y = "NDVI values", x = "Year") +
  facet_wrap(~hovedtype, ncol = 1)
```

@@

- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition. This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
- How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring. The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (Valøen, 2019). Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above. At the same time, wetlands are generally drier in the western parts of Norway than in the East (Moen et al. 2011), so NDVI patterns may, at least in part, be determined by moisture.

@@

**References introduced in the conclusion:**

Moen, A., Lyngstad, A. & Øien, D.-I. 2011. Kunnskapsstatus og innspill til faggrunnlag for oseanisk nedbørmyr som utvalgt naturtype. – NTNU Vitensk.mus. Rapp. bot. Ser. 2011-7: 1-72

Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. NTNU
